---
title: "Blog Post 4"
editor: visual
---

```{r}
#| echo: false
#| output: false
library(maps)
library(tidyverse)
library(readr)
library(here)
library(plotly)

state_df <- ggplot2::map_data("state")

cdc_maternal_mortality_rate_by_state <- read_csv(here("/Users/tessdowning/Desktop/MaternalMortality/posts/BlogPost4/CopyOfSYEdata/cdc_maternal_mortality_rate_by_state.csv"))
```

# Section 1: Mapping Maternal Mortality by State Rates

```{r}
#| echo: false
#| output: false
cdc_df <- cdc_maternal_mortality_rate_by_state |> mutate(state = str_to_lower(state))
cdc_full <- left_join(state_df, cdc_df, by = c("region" = "state"))
```

```{r}
#| echo: false
map2018 <- ggplot(data = cdc_full, aes(x = long, y = lat, group = group, label = `2018 rate (one-year)`,
                                       other_label = region)) +
  geom_polygon(colour = "black", aes(fill = `2018 rate (one-year)`)) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  scale_fill_distiller(type = "seq", direction = 1, palette = "Reds") +
  labs(title = "Maternal Mortality Rate (2018)",
       fill = "Mortality Rate")

ggplotly(map2018, tooltip = c("label", "other_label"))
```


In 2018, Arkansas had the highest rate of maternal mortality at 62%. Illinois had the lowest rate of maternal mortality with 15%.


```{r}
#| echo: false
map2019 <- ggplot(data = cdc_full, aes(x = long, y = lat, group = group, label = `2019 rate (one-year)`,
                                       other_label = region)) +
  geom_polygon(colour = "black", aes(fill = `2019 rate (one-year)`)) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  scale_fill_distiller(type = "seq", direction = 1, palette = "Reds") +
  labs(title = "Maternal Mortality Rate (2019)",
       fill = "Mortality Rate")

ggplotly(map2019, tooltip = c("label", "other_label"))
```

In 2019, Alabama had the highest rate of maternal mortality at 60%. Illinois had the lowest rate of maternal mortality with 14%. Arkansas had a rate of 38% which is not great but a large improvement from the previous years.

```{r}
#| echo: false
map2020 <- ggplot(data = cdc_full, aes(x = long, y = lat, group = group, label = `2020 rate (one-year)`,
                                       other_label = region)) +
  geom_polygon(colour = "black", aes(fill = `2020 rate (one-year)`)) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  scale_fill_distiller(type = "seq", direction = 1, palette = "Reds") +
  labs(title = "Maternal Mortality Rate (2020)",
       fill = "Mortality Rate")

ggplotly(map2020, tooltip = c("label", "other_label"))
```

Once COVID-19 hit in 2020, we can see a difference in the coloring of the map, centered around the South Eastern States. Tennessee had the highest maternal mortality rate at 69%. Pennsylvania had the lowest rate at 19% which is higher than the lowest states the previous two years. Arkansas is ranked second highest in this first pandemic year with 62% maternal mortality, almost double what it was in 2019. Alabama came in third with maternal mortality rate of 59%, one percent less than the year prior. Illinois saw an increase to 28% this year.

```{r}
#| echo: false
map2021 <- ggplot(data = cdc_full, aes(x = long, y = lat, group = group, label = `2021 rate (one-year)`,
                                       other_label = region)) +
  geom_polygon(colour = "black", aes(fill = `2021 rate (one-year)`)) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  scale_fill_distiller(type = "seq", direction = 1, palette = "Reds") +
  labs(title = "Maternal Mortality Rate (2021)",
       fill = "Mortality Rate")

ggplotly(map2021, tooltip = c("label", "other_label"))
```

2021 hit new highs all over the country for maternal mortality rates. Alabama is back to number one with a shocking maternal mortality rate of 86%. California had the lowest rate with 23%, the highest low value we have seen so far. Illinois rate crept up to 36% with Pennsylvania right behind it with 35%. Arkansas saw a rate of 67% and Tennessee had another shocking rate of 75%.
