---
title: "Blog Post 4: Bringing It All Together"
editor: visual
---

```{r}
#| echo: false
#| output: false
library(maps)
library(tidyverse)
library(readr)
library(here)
library(plotly)

state_df <- ggplot2::map_data("state")

cdc_maternal_mortality_rate_by_state <- read_csv(here("/Users/tessdowning/Desktop/MaternalMortality/posts/BlogPost4/CopyOfSYEdata/cdc_maternal_mortality_rate_by_state.csv"))
```



```{r}
#| echo: false
#| output: false
cdc_df <- cdc_maternal_mortality_rate_by_state |> mutate(state = str_to_lower(state))
cdc_full <- left_join(state_df, cdc_df, by = c("region" = "state"))
```

To better understand what the maternal mortality picture looked like during COVID-19, this last blog post will center around maps showing the different rates in each state. These maps were created using a group of CDC (Centers for Disease Control) data compiled and tidied by MuckRock. The dataset contains information about women of all ages in each state (MuckRock/Maternal-Mortality: A Repository of CDC National Center for Health Statistics Maternal Mortality Data and a Selection of State Data, from Michigan, Minnesota and North Carolina, about Maternal Deaths, n.d.). Plotted on each map are the maternal mortality rates, by state. These states were included in the dataset if there were more than nine pregnancy-related deaths in that year (MuckRock/Maternal-Mortality: A Repository of CDC National Center for Health Statistics Maternal Mortality Data and a Selection of State Data, from Michigan, Minnesota and North Carolina, about Maternal Deaths, n.d.).  That said, the CDC datasets contain valuable information from 24 states but are missing information from the other 26. Although we are missing this data, the maps still paint a good picture of the differences across the country. Another flaw of the data (and a flaw of our maternal health system) was that the data completely relied on death certificates. Because of the check box system described in Blog Post 1, the CDC’s only way to officially detect these deaths was if that little box was marked on the certificate. Because of this, there are some discrepancies if you were to compare the maternal death counts from this dataset to states’ reports. These maps were created using RStudio’s Quarto blog function and were the inspiration for this entire project! 
There are four maps covering the years 2018, 2019, 2020, and 2021.  The shade of red on each map represents the maternal mortality rate. The lighter the color, the lower the rate, and the darker, the higher the rate. Let us dive into the maps and the deeper meanings behind them. 

```{r}
#| echo: false
map2018 <- ggplot(data = cdc_full, aes(x = long, y = lat, group = group, label = `2018 rate (one-year)`,
                                       other_label = region)) +
  geom_polygon(colour = "black", aes(fill = `2018 rate (one-year)`)) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  scale_fill_distiller(type = "seq", direction = 1, palette = "Reds") +
  labs(title = "Maternal Mortality Rate (2018)",
       fill = "Mortality Rate")

ggplotly(map2018, tooltip = c("label", "other_label"))
```

In 2018, Arkansas had the highest rate of maternal mortality at 62. This same year, Illinois had a rate of 15.  Three years later in 2021, Arkansas released its first-ever review of maternal mortality. The fact that there is no previous, comprehensive report of the state’s maternal deaths could be a hint as to why Arkansas’ maternal mortality rate was the highest in the country for 2018. Without reporting previous statistics, it is hard to learn and gain knowledge from past problems. Some key findings in the Arkansas report revealed that “92% of pregnancy-related deaths were considered potentially preventable” (Brown and Canzoneri, 2021). This shocking statistic led the state to publish some potential solutions to the problem. These suggestions included providing contraception and offering tubal ligation to prevent unwanted pregnancy, implementing comprehensive safety bundles, and coordinated care within clinical settings, extending medical coverage for pregnant people to up to 60 days (about 2 months) or a year postpartum, expanding insurance coverages, and increasing medical access. The safety bundles help to standardize care by compiling materials needed to treat obstetric hemorrhages, hypertension, and drug-related complications in one spot (Brown and Canzoneri, 2021). By standardizing care, the treatment can be expedited to provide life-saving measures at a faster rate. As this blog will reveal in a few more paragraphs, these recommendations need to be implemented sooner. This delay in policy review and change could be a crucial factor in the (spoiler alert) increased rates of maternal mortality throughout the pandemic. Furthermore, in Arkansas, 37% percent of pregnancy-related deaths in 2018 occurred in Black women. In fact, for every white woman who died from a pregnancy-related death, more than two black women died (Brown and Canzoneri, 2021). As we know from the SDOH post in this blog, COVID-19 disproportionately affected women of color. Arkansas was setting itself up for failure with increased maternal deaths in the years to come. Older maternal age is like race as women who were 35+ had a higher maternal mortality ratio than their younger counterparts (Brown and Canzoneri, 2021). Again, as we previously learned the coronavirus became more severe with age. That said, 35 is not old, but there is a greater chance they will have more significant symptoms than those younger. There were comorbidities associated with these deaths including obesity, mental health conditions, and substance abuse (Brown and Canzoneri, 2021). These confounding problems are exacerbated by social determinants and could also be predictors for the increased rates we will see in 2020 and 2021. This would improve the checkbox system and provide us with more accurate numbers that could lead to better policies being created. Had the pandemic never happened, Arkansas could have been on its way to improving its statistics. This report was delayed by stay-at-home orders and the pause on the world, but I hope that the state continues to produce reports like this one. 

In 2018, Illinois released its maternal mortality report for 2015. This document can help us to under what steps the state took to achieve the lowest maternal mortality rate in the country. The state examined where there was potential for prevention in its system. It was revealed that there was poor communication between patients and providers, a lack of care coordination, hospitals lacked clear policies for the identification and treatment of both pregnant and postpartum women, there were untaken opportunities for screening, inadequate provider skill, and knowledge, providers did not know where to refer women who were struggling with mental health or substances, and there was an overall lack of public knowledge about proper postpartum care (Barker et al., 2018). These are huge holes in maternal health policy but are impressive to acknowledge through a multifaceted and holistic approach to maternal mortality research. By recognizing these problems, Illinois was able to suggest some solutions to improve its pregnancy-related death rates by targeting both access and quality of health care, comorbidities, and SDOHs. There were suggestions aimed at hospitals, insurance companies, friends and families, other medical providers, and to the State of Illinois. The recommendations for the government included ensuring women who are at different risks of complications are brought to the appropriate level medical center, expanding access to doulas and midwives to target and aid high-risk mothers, increasing funding to educational sources for information on maternal mortality, increasing access to substance use and mental health services targeted for women and their children, and to conduct postpartum depression screenings for mothers during their baby’s doctor appointments in the first year of life (Barker et al., 2018). These changes in 2015, led to Illinois success in 2018 and their recommendations continued to work through 2019. 

```{r}
#| echo: false
map2019 <- ggplot(data = cdc_full, aes(x = long, y = lat, group = group, label = `2019 rate (one-year)`,
                                       other_label = region)) +
  geom_polygon(colour = "black", aes(fill = `2019 rate (one-year)`)) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  scale_fill_distiller(type = "seq", direction = 1, palette = "Reds") +
  labs(title = "Maternal Mortality Rate (2019)",
       fill = "Mortality Rate")

ggplotly(map2019, tooltip = c("label", "other_label"))
```

Alabama had the highest rate of maternal mortality, 60, in 2019. In 2018 and 2019, Alabama’s biggest pregnancy-related death troubles were infections, cardiovascular events, and hemorrhaging and these problems disproportionately affected black, non-Hispanic women (Alabama Maternal Mortality Review Program | Alabama Department of Public Health (ADPH), n.d.). For all races, as age increased, pregnancy-related deaths also increased with the highest numbers being for those in the 30+ demographic (Alabama Maternal Mortality Review Program | Alabama Department of Public Health (ADPH), n.d.). These also relate to the problems Arkansas had with health disparities. As we learned earlier in the blog, women of color are disproportionately at risk of maternal mortality because of systematic racism (Villarosa, 2018). Some of the recommendations made to aid these issues include increasing access to mental health and substance use services, providing implicit bias training to all medical staff, enhancing care coordination, educating women with high-risk conditions on the risks they could experience while pregnant, providing monitoring for those at high risk of cardiovascular events, and standardizing protocol (Alabama Maternal Mortality Review Program | Alabama Department of Public Health (ADPH), n.d.). Illinois still had the lowest rate of 14 and Arkansas saw an improvement when their rate dropped to 38. 

```{r}
#| echo: false
map2020 <- ggplot(data = cdc_full, aes(x = long, y = lat, group = group, label = `2020 rate (one-year)`,
                                       other_label = region)) +
  geom_polygon(colour = "black", aes(fill = `2020 rate (one-year)`)) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  scale_fill_distiller(type = "seq", direction = 1, palette = "Reds") +
  labs(title = "Maternal Mortality Rate (2020)",
       fill = "Mortality Rate")

ggplotly(map2020, tooltip = c("label", "other_label"))
```

Once the coronavirus hit in 2020, we can see the largest differences in the map thus far. Most of the dark red coloring is in the Southeastern States. Tennessee had the highest rate in 2020 of 69. Following closely behind, Arkansas was in second with 62 and Alabama was third with a rate of 59, almost double what it was the year prior. Tennessee’s COVID-19 policies could have something to do with this. To compound the impact of COVID, in early March 2020, Tennessee saw one of the deadliest tornados in its history, and the damage from this natural disaster certainly did not help the state’s maternal mortality rates for that year. Despite the billions of dollars in damages, TN did assemble a COVID-19 task force on March 4th (COVID-19 Timeline, n.d.). However, instead of hard cutoffs for closing schools, working from home, and attending large events, Tennessee advised its citizens to use discretion.  Tennessee had access to COVID-19 tests early and distributed them along with emergency grants to its citizens. The state began to reopen in late April after devasting losses from the disease and more natural disasters (COVID-19 Timeline, n.d.). With these large-scale natural disasters comes a lot of uncertainty in many aspects of life including housing, money, work, support systems, food supply, water supply, and more. The combination of optional COVID-19 policies and the disaster-caused disadvantages on the SDOHs contributed to the rates of maternal mortality in Tennessee. 

Pennsylvania had the lowest rate in 2020 of 19 maternal deaths while Illinois saw a jump to 28. Pennsylvania was quick to jump into action when the pandemic hit, starting to prepare in early February. On that fateful March 16th, PA closed restaurants, imposed no visitors rules at nursing homes and prisons, closed schools, and childcare centers, and more (“Pennsylvania - Coronavirus State Actions,” 2020). As the pandemic progressed, they remained vigilant following “Stay at Home” orders, distributing masks, and shutting down xenophobia. The Governor also signed many executive orders that protected the people of PA from eviction, foreclosures, malnourishment, and more (“Pennsylvania - Coronavirus State Actions,” 2020). When it comes to carrying a healthy baby to term and delivering it, it is important to reduce stress, have a safe place to live, and have nutritious foods to eat. Because the PA government protected these rights, their maternal mortality rates did not soar to the highs of other states. 

```{r}
#| echo: false
map2021 <- ggplot(data = cdc_full, aes(x = long, y = lat, group = group, label = `2021 rate (one-year)`,
                                       other_label = region)) +
  geom_polygon(colour = "black", aes(fill = `2021 rate (one-year)`)) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  scale_fill_distiller(type = "seq", direction = 1, palette = "Reds") +
  labs(title = "Maternal Mortality Rate (2021)",
       fill = "Mortality Rate")

ggplotly(map2021, tooltip = c("label", "other_label"))
```

In 2021, there were new highs all over the country in terms of maternal mortality. Alabama had the highest rate that year, a shocking 86. Alabama confirmed its first case of the coronavirus on March 13, 2020, declared a state of emergency, and closed all its schools for three weeks on the same day (Fiscus, 2020). As the year went on and cases were growing to new highs, Alabama began to reopen with retail stores, gyms, and hairdressers. This led to Montgomery County becoming a hot spot for the entire country as its cases had doubled in only two weeks (Fiscus, 2020). Alabama started vaccinating in early 2021 for those 75 or older and eventually rolled out vaccines to all age demographics (Documenting Alabama’s Path to Recovery from the Coronavirus (COVID-19) Pandemic, 2020-2021, n.d.). On April 9th, 2021, the governor lifted major restrictions on businesses and individuals including the statewide mask requirement. Following this, in May, the state announced that it would end all federal benefits by the end of June (Documenting Alabama’s Path to Recovery from the Coronavirus (COVID-19) Pandemic, 2020-2021, n.d.). Alabama was extremely focused on reopening which caused an uptick in cases. This drive to reopen most likely also affected people’s financial situations. With the economy unstable and Alabama ending its benefits, many expecting parents were probably having a challenging time making ends meet after their jobs were affected by a pandemic for a year. Arkansas had a rate of 67 in 2021 and Tennessee had some improvement but still a shocking rate of 75. 

California had the lowest maternal mortality rate of 23, which is the “highest” low out of the four years. California was early to begin COVID restrictions with events being canceled as early as January 28th (Fiscus, 2020). These were Lunar New Year events, so they disproportionality targeted specific populations, however, this officially kicked off California’s strong COVID-19 prevention strategies. The state’s state of emergency and COVID-19 taskforce was created on February 26th, the first in the US (Fiscus, 2020). In early 2021, California was the first state to exceed 3 million cases and over 30,000 deaths, and many of the early restrictions like travel bans and masks were brought back (Fiscus, 2020). As the year went on California became very vigilant and efficient with their vaccinations and made sure that disadvantaged communities had the opportunity to be vaccinated (Fiscus, 2020). As all of this was going on the state was protecting its homeowners and helping local businesses stay afloat by altering laws, such as bringing alcoholic beverages to go (Fiscus, 2020). Liquor sales are a major contributor to a restaurant’s earnings. By providing businesses with the opportunity to sell alcohol to go, they could still generate a portion of the profits they were earning before the pandemic. This also provided homeowners and those isolated by stay-at-home orders with a small sense of normalcy. California was very on top of stopping the spread and keeping the community safe. Due to this, they mitigated the increased risks that other mothers were facing around the country. Illinois’ rate crept up to 36 and Pennsylvania was right behind it in the data with a rate of 35. 
These highlighted states are examples of the steps governments should and should not take when trying to protect the lives of their expecting mothers. The states that were actively pursuing improvements to their maternal health outcomes before the pandemic, were the more prepared when COVID-19 hit. The southern states like Arkansas, Alabama, and Tennessee were already struggling, and the coronavirus took advantage of the cracks in their systems. The main lesson states learned from the pandemic was the importance of having a set protocol. Most hospitals were disorganized and did not have regulated plans for when women fell into emergencies. By creating streamlined, universal plans, medical providers can quickly provide the life-saving care mothers and their children need. 

After taking the time to read these four posts, I hope the readers of this blog were able to gain a better sense of how maternal mortality is measured in the US. These records help us to visualize how the impact of social determinants of health and comorbidities played into these rates throughout the COVID-19 pandemic. If you have any questions about anything written in these blogs, please feel free to contact me via the links in the About section. Thank you so much for taking the time to read these posts. 






